Type Evolution in Programming Systems

### 2024-05-23 paper.tex
Edits in intro, update arrow sketches

\usepackage{inputenx}
\usepackage[backend=biber]{biblatex}
\addbibresource{paper.bib}

@@

\title{Schema Change: Challenge Problems}

@@

@@

@@

\newpage
\section{Dimensions of change}

\begin{figure}[t]
\vspace{-1em}
\centering
\includegraphics[width=0.5\textwidth]{figures/layers.png}
\caption{Pace layers from Stewart Brand's classic \emph{How Buildings Learn} \cite{Brand95}.
  The outer layers rarely change, while the inner layers change frequently. The geographic
  site of a building is eternal, its structure remains stable for decades, space plan
  needs to change every couple of years. ``Because of the different rates of change of its
  components, a building is always tearing itself apart.''
  \textcolor{red}{[JE] Except the outer skin changes faster than the structure.}
  }
\label{fig:layers}
\end{figure}

\textcolor{red}{[JE] `change in a program' might be misunderstood as changes to the source code. Perhaps `system' or `software system' might be better than `program'? }

In the broadest sense, this paper is concerned with the difficulties posed by a change
in a program. The first dimension of difficulties that we consider in this paper is
local. When something in a program changes, other parts of the program typically
need to be brought to sync with the changed part. The complexity of the problem varies.
If nothing else in the program depends on the changed part, for example when changing
a runtime value of an object field, the change poses no difficulty.
If many other parts of the program depend on the changed part, the difficulty is greater.
For example, if we change the schema of data structures used in the program, we also need to
change code that works with it and existing data that the users created. To structure our
discussion about this part of the problem, we will use the idea of pace layers introduced by
Brand \cite{Brand95,Brand18} (Figure~\ref{fig:layers}).

The second dimension that we consider in this paper is non-local. A program may have multiple
variants that exist independently and need to be synchronized. The difficulty of the problem
depends on what kind of synchronization we want to support. In the easier case, all variants
of the program converge, i.e.~all variants should eventually adopt all changes. In the more
challenging case, variants of a program may diverge, i.e.~a user may want to apply only certain
changes from other users. Note the non-local dimension may interact with the pace layers of the
local dimension. A typical case may be that the code and schema of a program should converge,
while data and the transient state of a program may diverge.

@@

@@

Schema evolution has long been a major problem for SQL databases, which provide suprisingly little help. SQLite~\cite{sqliteDatatypes} uses dynamically typed values allowing them to be implicitly converted if the datatype of the column changes. MYSql~\cite{mysqlAlterTable} can reorder columns without destroying their data. But apart from such special cases, SQL databases have no general purpose support for data migration. As a result in practice schema evolution is still mostly done by writing custom SQL code to alter the schema and migrate the data. Such custom code is greatly complicated if it needs to be done without taking the database offline or must be coordinated across multiple shards. In \textit{Refactoring Databases} \citet{ambler06} offer a comprehensive taxonomy of schema evolution patterns, including typical strategies for online migration and sample SQL implementations.

@@

@@

>>>>

\usepackage{inputenx}
\usepackage[backend=bibtex]{biblatex}
\addbibresource{paper.bib}

@@

\title{Type Evolution in Programming Systems}

@@

Many past improvements in the practice of programming worked by speeding up feedback loops,
both in collaboration with other programmers (as in collaborative development tools
\cite{goldman2011real,kurniawan2015coder,replit}) and in individual's interaction with the
programming environment (as in live programming \cite{rein2018exploratory}).

To further tighten the feedback loop and make programming more effective, future programming
systems will need to enable programmers to collaborate at a fine-grained level, letting them
adopt individual changes created by their collaborators (as in local-first software
\cite{localfirst}). To provide immediate feedback, future programming systems will need to
transcend treating state as ephemeral, tied to the lifetime of a running program and reset when the
program is edited.

We need to shift our focus to \emph{programming systems} \cite{techdims} where program state
is persistent and co-evolves with the behavior and data of the system. This is already the case
in collaborative applications with state persistent in a database, image-based systems like
Smalltalk and live programming environments with hot-reloading where the program can be edited
without resetting all state.

While we are firmly convinced that more collaborative and live programming systems are the
future of programming, there are a number of problems that need to be resolved in order for
such systems to become a reality. The key problem is of \emph{type evolution}, i.e., how to
handle the change of schema of the data program works with in a way that preserves integrity
constraints, existing data and the logic of code.

Type evolution is not a new problem. It exists in various forms in a range of existing systems.
In the database world, the problem of schema evolution is well-studied \cite{erhard06}.
It typically focuses on ensuring data integrity, although co-evolution of database structure
and code that manipulates the database has also been studied \cite{Cleve2006,wang2019synthesizing}.
In live programming, the problem is known as hot code reloading \cite{barenz2020essence,beckmann2021shortening},
although it typically focuses on updating transient state and ignores the problem of more complex
data migrations. The problem is equally studied in the bidirectional transformation (BX)
communities \cite{czarnecki2009bidirectional} and in work on model-driven development
\cite{metamodels,alanen2003difference}.

Type evolution is an important challenge for future programming systems. But looking at the problem
from the perspective of programming systems can also offer a new perspective that
unfies the many different aspects of the problem partially explored in other areas. A programming
system may have permanent data (as in a database) whose structure can change, requiring
corresponding change to data and code. A programming system may have a transient state conforming
to a type structure that may change during live programming. A programming system may also need
to allow programmers to independently work on different versions of a program.

In this paper, we draw connections between these different kinds of type evolution, pointing out
essential shared structure. By doing so, we aim to motivate further research on this important
problem and encouraging knowledge sharing across subfields. Our contributions are twofold.
First, we present a shared conceptual framework for thinking about the relationship between
schema, code, and data in a programming system. Second, we present five case studies that use
the framework to study type evolution in a highly diverse set of real-world programming scenarios.

\begin{figure}[t]
\centering
\vspace{-1em}
\includegraphics[width=0.45\textwidth]{figures/layers.png}
\caption{Pace layers from Stewart Brand's classic \emph{How Buildings Learn} \cite{Brand95}.
  Different layers of a building change at different pace. The geographic
  site of a building is eternal, its structure remains stable for decades, space plan
  changes every couple of years.}
\label{fig:layers}
\end{figure}

\section{Type Evolution}
To paraphrase Stewart Brand \cite{Brand95}, ``almost no programs evolve well. But all programs
(except for monuments) evolve anyway, however poorly, because the usages in and around them are
changing constantly.'' Programs evolve both while they are being created and during their operation
in response to new needs. As with buildings (Figure~\ref{fig:layers}), different aspects of
program evolve at different pace. And ``because of the different rates of change of its components,
a program is always tearing itself apart.''

\subsection{Types, Code and Data}

As is the case for buildings, making a change at some layers (skin or stuff) is relatively easy,
but a change at a more basic layer (structure) has cascading effects on layers that depend on it.
In the case of programs, a more basic layer is a structure that defines the shape of data and code.
We refer to this structure as \emph{types}, although we do not use the term in an precise technical
sense. In our conceptual framework, types may be data types that define the shape of data. They
may be explicitly declared in code and checked statically, as well as dynamically checked or
even implicit. The key property of types for us is that they define the shape of other layers
of a program, such as data and code. As a result, a change of program types requires a
corresponding change in program data and code.

The situation is illustrated in Figure~\ref{fig:change} using two variations of a diagram that
we will return to repeatedly in this paper. When a type changes from $T_1$ to $T_2$,
we need to find and apply a corresponding change to other parts of the program whose shape
is defined by the changed type. The problem of \emph{type evolution} is thus finding and
applying changes that will update data $D_1$ and code $C_1$ into new versions $D_2$ and $C_2$
that correspond to the new type.

The complexity of the problem varies. The first dimension of challenges that we consider
is local. The complexity of the challenge depends on how many other parts of the program
depend on the modified type. If both code and data depend on the type, the complexity of the
program is greater. If there is no associated code or the existing data can be safely discarded,
the complexity is lesser.

The second dimension of challenges that we consider is non-local. A program may have multiple
variants that exist independently and need to be synchronized. The difficulty of the problem
depends on what kind of synchronization we want to support. If all variants of the program should
eventually adopt all changes (converge), the complexity is lesser. If we allow user to apply
only certain changes from other users (divergence), the complexity is greater. A typical case
may be that the types and schema of a program should converge, but data may diverge.

\begin{figure}[t]
\centering
\includegraphics[width=0.65\textwidth]{figures/arrows.png}
\caption{When a type changes from $T_1$ to $T_2$, a corresponding change needs to take place
  in data and code whose shape is defined by type. The challenge is finding a suitable
  corresponding transformation (semi-)automatically.}
\label{fig:change}
\end{figure}

\subsection{Conceptual Framework}

@@

\textcolor{red}{TODO: got as far as here so far...}

@@

@@

% elm
% database
% divergence
% cambria
% documents
% state machine

% DIVERGENCE IN TRANSITION (because you cannot do big bang update) c.f. Tijs' example

% typically people just blow away the state but Elm actually has demos
% of people doing hot reload without losing state.
% (maybe they never change the type?)

% ELM EXAMPLE
% current example ("completes") is contrived - it is equivalent data
% more interesting is when you lose / add data
% add "completed" with default instead

@@

TODO: Maybe insert something about methodology in general?
(ref the table challenges paper? although this paper is no longer so much about challenges)

\paragraph{Schema evolution.}
Schema evolution has long been a major problem for SQL databases, which provide suprisingly little help. SQLite~\cite{sqliteDatatypes} uses dynamically typed values allowing them to be implicitly converted if the datatype of the column changes. MYSql~\cite{mysqlAlterTable} can reorder columns without destroying their data. But apart from such special cases, SQL databases have no general purpose support for data migration. As a result in practice schema evolution is still mostly done by writing custom SQL code to alter the schema and migrate the data. Such custom code is greatly complicated if it needs to be done without taking the database offline or must be coordinated across multiple shards. In \textit{Refactoring Databases} \citet{ambler06} offer a comprehensive taxonomy of schema evolution patterns, including typical strategies for online migration and sample SQL implementations.

@@

\paragraph{Live, MDD, etc. programming?}

% Some references from Tomas
%
% collaborative programmign tools
\cite{goldman2011real,kurniawan2015coder,replit}
% exploratory/live review paper
\cite{rein2018exploratory}
% hot reload in live systems
\cite{barenz2020essence,beckmann2021shortening}

% Some references from Tijs
%
% SQL+code
\cite{Cleve2006,wang2019synthesizing}
% BX
\cite{czarnecki2009bidirectional}
% MDD
\cite{metamodels,alanen2003difference}.

@@

\appendix
